---
subtitle: "Minianalysis 2"
title: "Predicting car mileage"
author: "Joel Goop"
date: \today
dev: svg
output:
  revealjs::revealjs_presentation:
    theme: serif
    highlight: pygments
    transition: slide
    css: custom.css
    self_contained: false
    center: false
    fig_caption: true
---
```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)

load('Lab1data.RData')
load('itest.RData')
Auto.Train<-newAuto[-itest,]
Auto.Test<-newAuto[itest,]

source('funcs.R')

source('imputation.R')
```

## Pre-processing and assumptions

- Inverse transform mileage `mpg` $\rightarrow$ `gpm`
- Imputation using regression (works well for highly correlated variables)
- Model selection using both exhaustive search and backward elimination

## Mean squared errors

- Lowest prediction MSE is 5.66
- Without imputation increases to 6.13
- Prediction errors lower than training?
- Backward elimination stops to soon!

```{r mses,dev='svg',echo=FALSE, results="hide", fig.width=7, fig.height=4}

source('regsubsets_test.R')
p
```


## Model

- Best predictive model contains only `weight` and `year`
- Prediction MSEs of larger models are close
     - Additional variables do explain more of `mpg`
     - Not enough to justify increasing $V\left(\hat{\beta}\right)$

```{r model}
m <- lm(gpm ~ weight+year, data=imp.auto.train)
```

```{r modelsummary,echo=FALSE}
s <- summary(m)
knitr::kable(s$coefficients)
```
